---
title: "Reproducible Research: Peer Assessment 1"
author: "Anna Popova"
date: "1 Jun 2015"
output:
  html_document:
    keep_md: true
---

## Introduction
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Data
The data for this assignment can be downloaded from the course web site.

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD     format

* **interval**: Identifier for the 5-minute interval in which     measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

At first, we set locale from Russian to English
```{r set_locale}
Sys.setlocale("LC_TIME", "English")
```

Then we get file with data from URL and unzip it
```{r get_file}
FileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

if (!file.exists("repdata-data-activity.zip")) {
   download.file(FileUrl, destfile = "repdata-data-activity.zip")
}
unzip("repdata-data-activity.zip")
```

## Assignment

### Loading and preprocessing the data

Load the data to `activity_data`
```{r load_data}
activity_data <-
    read.csv('activity.csv', header = T)
```

Transform the data into a format suitable for your analysis
```{r transform_data}
activity_data$date <- 
    as.Date(activity_data$date)
```

### What is mean total number of steps taken per day?
For this part of the assignment we ignore the missing values in the dataset.

#### Make a histogram of the total number of steps taken each day

Summarize the total number of steps taken by each day
```{r make_hist}
StepsByDay <-
   aggregate(x = activity_data$steps,
             by = list(activity_data$date),
             FUN = sum,
             na.rm = T)
names(StepsByDay) <- c("date","steps")

hist(StepsByDay$steps, 
     breaks = 20,
     main = "Histogram of steps taken each day",
     xlab = 'Total Number of Steps',
     col = 'grey')
```

#### Calculate and report the **mean** and **median** total number of steps taken per day
Calculate mean and median total number of steps taken per day
```{r mean_calc}
mean_steps = round(mean(StepsByDay$steps), 2)
median_steps = round(median(StepsByDay$steps), 2)
```

draw lines for mean and median
```{r make_lines}
abline(v = mean_steps, lwd = 3, col = 'blue')
abline(v = median_steps, lwd = 3, col = 'red')
```

place legend
```{r make_legend}
legend('topright',
       lty = 1,
       lwd = 3,
       col = c("blue", "red"),
       cex = .8, 
       legend = c(paste('Mean: ', mean_steps),
       paste('Median: ', median_steps))
       )
```

### What is the average daily activity pattern?
#### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)



#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


### Imputing missing values
There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

#### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

#### Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

#### Create a new dataset that is equal to the original dataset but with the missing data filled in.

#### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?


### Are there differences in activity patterns between weekdays and weekends?
